---
title: "A Quick Tour through State Space Reconstruction"
subtitle: "Short Course 'An Introduction to Empirical Dynamics Modelling, ICTP SAIRF', Tutorial #1"
author: "Brenno Cabella, Paulo Inácio Prado, Renato Coutinho,  Marina Rillo, Rafael Lopes, Roberto Kraenkel"
date: "ICTP SAIFR, School on Physics Applications in Biology, January 2018"
output:
         rmdformats::readthedown:
         self_contained: true
         thumbnails: true
         lightbox: true
         gallery: false
         highlight: tango
         toc_depth: 4
---

```{r setup, echo=FALSE, warning=FALSE, message=F}
library(knitr)#; library(zoo); library(xts)
library(rEDM)
library(ggplot2)#; library(cowplot)
library(dplyr); library(tidyr)
#library(deSolve)
library(xts)
library(rgl)
library(plotly)
source("../utilities/make_block.R")

opts_chunk$set(fig.align = 'center',
               fig.show = 'hold',
               fig.height = 5,
               fig.width = 5, 
               warning = FALSE, message = FALSE, error = FALSE, echo=TRUE)
options(formatR.arrow = TRUE,width = 90)###, cache=TRUE)
```


This tutorial presents the general idea of atractor reconstruction with syntetic data, 
and then we propose exercises with simulated and real data analysis proposed.
For these activities, you will need the most recent version of
[R](https://cran.r-project.org/) 
and the rEDM package installed in your work computer. 
To start, open R and load the package by typing the follwoing code in the R console:


```{r load rEDM, echo=TRUE}
library(rEDM)
```


Before you proceed
please read pass through the essential concepts by 
reading the introduction and the first section (*"Empirical Dynamic Modelling"*) of
rEDM's tutorial that you can find [here](https://cran.r-project.org/web/packages/rEDM/vignettes/rEDM-tutorial.html) 
or in your local R installation with the command


```{r call vignette, eval=FALSE}
vignette("rEDM-tutorial", package="rEDM")
```

# A simple case: harmonic oscilator

This first part is to get the grips with R language and to figure out how 
to reconstruct a state space with only one time series and its lags.
For the first example, we use a very well known system and its representation on state space, 
the bidimensional harmonic oscilator:


$\frac{dx}{dt}=y$

$\frac{dy}{dt}=-x$

which has the solution $x(t)=\sin(t), \ y(t)=\cos(t)$ and so we can draw the state 
space of this system by laying down in two  cartesian axis its two degrees of
freedom, $x(t)$ and $y(t)$. This system has a circular atractor, that
is, all the possible states lie over a circle in the state space:

```{r generate data for the harmonic oscilator, echo=TRUE}
## Generate the data at tau = 0.2 time intervals
time1<-seq(0, 500, by = 0.2) 
x1<-sin(time1)
y1<-cos(time1)
## Plot the state space
plot(x1, y1, type = 'l', main = "State-space of harmonic oscilator",
     cex=0.2, xlab="x(t)", ylab="y(t)")
```

And here are the time series for the two state variables:

```{r time series of each variable}
## Time series of the 1st 250 observations
ind <- 1:250
plot(time1[ind], x1[ind], type = 'l',
     main = "Time series of the variables of the system",
     col = "red", xlab = "Time", ylab = "Amplitude",
     ylim=c(-1,1.25))
lines(time1[ind], y1[ind], col = "blue")
legend("topright", c("x","y"), col=c("red", "blue"),
       lty=1, bty="n")
```

## A reconstruction of the state space

Now suppose that all you information you have about this system is 
a time series of one of the state variables, say $x$. 
Because $y(t)$ is missing we are not able to construct a state
space as we saw above, but by using the result of Takens theorem we can
construct a **shadow manifold** that recovers the information of 
the true state space.  This reconstruction is not a copy of 
the original atractor, but it is a valid topological reconstruction of the
original, a "shadow". 


To do this, we 
generate new variables by lagging the time series 
by multiples of a time interval $\tau$. In this case we will use
the interval between observations ($\tau=0.2$ time units).
The R commands below creates a data table object ( which is called  a
*data frame* in R) with the original time series and its lagged
version. The  handy function `make_block` by the rEDM authors does this
job. Use this [link](../utilities/make_block.R) to download the function to your working directory
and then load it with the command:

```{r load make_block, eval=FALSE}
source("../utilities/make_block.R")
```

And now you can create a lagged variable dataframe in R by copying the following code and
pasting in the R console:


```{r plot one time series alone}
## Data frame with observations with tau = 0.2 with one lag
df1 <- make_block(data.frame(x=x1), t=time1, max_lag = 2)
```

A quick look at the first and last lines of the data frame
can help to understand the lagged variabel structure:


```{r check lagged dataframe}
## Take a look in the lagged data to understand how the function works
head(df1)
tail(df1)
```

By ploting the time series
as a function of their lagged versions we get the
shadow manifold, which is similar to the original one: 

```{r shadow manifold}
plot(x_1 ~ x, data=df1, type="l",
     xlab = "X", ylab = "X(t + tau)" )
```

There is a minimum number of dimensions to build
a valid rceonstruction of the original atractor.
In this simple case we obviously need two dimensions.
Takens Theorem guarantees that the number of dimensions
is up to 2D + 1, where D is the dimmensionality of the 
the attractor. Finding the optimal dimension for
the shadow manifold is part of a vast topic called *Embedology*
(Sauer *et al.* 1991). The package `rEDM` 
has some methods to find
the optimal dimensions of the manifold. 
We will learn more in the following tutorials of this course.


# Noisy data

How robust is the atractor reconstruction to random noise?
There are two main types of noise or error in data, which we call
**process errors** and
**observational errors**. The first one is about random changes of the
state variables, like environmental fluctuations that affect the
dynamic. The
second type of error appears when we are not able to measure
the data with total precision, when we miss some counts on the counting of
individuals of species, or do error in the sampling or measurements.
Observational errors do not affect the dynamic, only the portrait
we make of it.

In this section, we check how robust the atractor is to error by adding simulated noise to the original time series.
Use the commands below to
simulate that the time series has independent measurement errors with constant variance.
To do this, we add to each observation a value drawn from a Gaussian distribution with zero mean and
standard deviation which is 1/12 of the standard deviation of the
time series itself. 


```{r plot one 2nd time series alone}
## a new vector of observed values + Gaussian iid error
x2 <- x1 + rnorm(n = length(x1), mean = 0, sd = sd(x1)/12)
## Time series
plot(x1[ind] ~ time1[ind], type="l", xlab = "X", ylab = "Time",
     col="grey", lwd = 5)
lines(x2[ind] ~ time1[ind], type="l", col="red", lwd=2)
```

We thus proceed by creating a new data frame with the lagged variables 
and by ploting the new shadow manifold:

```{r 2D plot noisy series}
## Data frame with lagged variables
df2 <- make_block( data.frame(x = x2))
## Shadow manifold with embedding = 2
plot(x_1 ~ x, data=df2, type="l",
     ylab = "X(t + tau)", xlab = "X" , col="grey",
     lwd=0.5)
```

In this case the shadow manifold still looks like the 
original shape of the attractor, despite the noise caused by measurement errors.

## Increasing the noise

Measurement errors make atractors more complex by spreading
points (and thus trajectories) over the state space. 
Hence, the trajectories may cross in the shadow manifold,
which breaks down the one to one  mapping to the original atractor.
Of course this effect gets larger as we increase the measurement
error. We now will investigate a bit more how this affects
the atractor reconstruction.
The commands below doubles the measurement error of the time series
and plots the shadow manifold:

```{r more noise}
x3 <- x1 + rnorm(n = length(x1), mean = 0, sd = sd(x1)/6)
df3 <- make_block(data.frame(x=x3), max_lag = 5)
plot(x_1 ~ x, data=df3, type="l", col="grey", lwd=0.5,
     ylab = "X(t + tau)", xlab = "X" )
```


The atractor reconstruction looks worse now, but we can overcome this
problem by adding more dimensions to the shadow manifold. 
By doing this we unfold the shadow manifold, that is
unfolding crossing trajectories.
We do this adding more axes, that correspond to 
the time series lagged by the further time steps 
($t + \tau, \ t + 2\tau, \ t + 3\tau$...).
Even the if the original manifold has a lower dimension its is valid to
reconstruct the state space with more dimensions[^1]:


You can check with the interactive plot below
how an additional dimension unfolds a bit more the reconstructed
atractor.  In the following days we will learn some methods to find
the optimal dimensions of the manifold. By now the take-home message
is: the more complex the atractor is, the more dimensions you
will need to have a good reconstruction.


```{r more noise 3D plot, echo=FALSE}
plot_ly(df3, x = ~x, y=~x_1, z=~x_2, type="scatter3d",
        mode="lines", opacity=0.25) %>%
    layout(scene=list(camera = list(eye = list(x = 0, y = 0, z = -2)),
                      xaxis = list(title = 'X'),
                      yaxis = list(title = 'X (t + tau)'),
                      zaxis = list(title = 'X (t + 2tau)')))
```

# Problems

## What is the effect of $\tau$ ?

The commands below simulate a new time series of the bimensional
harmonic oscilator without measurement
error but with a much smaller interval between observations 
($\tau=0.01$ time units). 


```{r smaller tau}
## Generate the data at 0.01 time intervals
time4<-seq(0, 500, by = 0.01) 
x4<-sin(time4)
y4<-cos(time4)
```

This change in $\tau$ affects the shape of the reconstruction:

```{r smaller tau manifold}
df4 <- make_block(data.frame(x=x4), t=time4)
plot(x_1 ~ x, data=df4, type="l",
     xlab = "X", ylab = "X(t + tau)" )
```

How do you explain that? Which practical issues does this fact bring
and how to solve them?

### Hints

**1.** The command above adds a line $x(t) = x(t + tau)$ to the manifold, which
can help you to understand the problem with small values of $\tau$:

```{r abline, eval=FALSE}
abline(0,1, col="red")
```

**2.** Check this site: (http://www.physics.emory.edu/faculty/weeks//research/tseries4.html)


## Atractor reconstruction

By using only the functions learned at this tutorial, try to
reconstruct the state space of the time series] in this [csv
file]("problem-time-series.csv"), 
checking if a three-dimensional manifold improve the reconstruction.
To load the data in R download the csv file and run the command:

```{r problem time series}
prob2 <- read.csv("problem-time-series.csv")
```
And to take look in the time series plot run:

```{r plot time series problem 2}
plot( X ~ time, data = prob2, type="l")
```

### Hints

**1.** Keep in mind the insight from the previous problem: 
the value of $\tau$ 
change the shape of the shadow manifold.

**2.** To create a simple 3D plot in R you can use the 
`scatterplot3d` package. The example below 
does this for one of the noisy
timeseries we examined above:

```{r 3D plot with scatterplot3d}
## Loads the package, if not available
library(scatterplot3d)
scatterplot3d(x = df2$x, y = df2$x_1, z = df2$x_2,
              type = "l", color="grey", lwd = 0.5,
              xlab = "x(t)", ylab = "x(t+tau)", zlab = "x(t + 2 tau)")
```

If the package is not available in your local R installation
you can install it with:


```{r install packages, eval=FALSE}
install.packages("scatterplot3d")
```


# Learn more

* Sugihara, G., May, R., Ye, H., Hsieh, C.H., Deyle, E., Fogarty,
  M. and Munch, S., 2012. Detecting causality in complex
  ecosystems. science, 338(6106), pp.496-500.
* DeAngelis, D.L. and Yurek, S., 2015. Equation-free modeling unravels
  the behavior of complex ecological systems. Proceedings of the
  National Academy of Sciences, 112(13), pp.3856-3857. 
* [Taken's Theorem](https://www.youtube.com/watch?v=QQwtrWBwxQg&list=PL-SSmlAMhY3bnogGTe2tf7hpWpl508pZZ&index=2), 
a video by Sugihara's Lab.
* Takens, Floris. Detecting strange attractors in turbulence. Lecture notes in mathematics 898.1 (1981): 366-381.
* Deyle, E.R. and Sugihara, G., 2011. Generalized theorems for
  nonlinear state space reconstruction. PLoS One, 6(3), p.e18295. 
* Sauer, T., Yorke, J.A. and Casdagli, M., 1991. Embedology. Journal
  of statistical Physics, 65(3), pp.579-616. 


[^1] Sauer, Tim, James A. Yorke, and Martin Casdagli. "Embedology." Journal of statistical Physics 65.3 (1991): 579-616.


